---
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Probability and Statistics

# Lab Assignment 1: Naive Bayes Classifier

## Work breakdown

-   *Name1 Surname1*:
-   *Name2 Surname2*:
-   *Name3 Surname3*:

## Introduction

During the first three weeks, you learned a couple of essential notions
and theorems, and one of the most important among them is the **Bayes
theorem**.

**Naive Bayes Classifier** is a simple algorithm, which is based on
**Bayes theorem** and used for solving classification problems.
**Classification problem** is a problem in which an observation has to
be classified in one of the $n$ classes based on its similarity with
observations in each class.

It is a **probabilistic classifier**, which means it predicts based on
the probability of an observation belonging to each class. To compute
it, this algorithm uses **Bayes' formula,** that you probably already
came across in **Lesson 3:**
$$\mathsf{P}(\mathrm{class}\mid \mathrm{observation})=\frac{\mathsf{P}(\mathrm{observation}\mid\mathrm{class})\mathsf{P}(\mathrm{class})}{\mathsf{P}(\mathrm{observation})}$$

Under the strong **independence** assumption, one can calculate
$\mathsf{P}(\mathrm{observation} \mid \mathrm{class})$ as
$$\mathsf{P}(\mathrm{observation}) = \prod_{i=1}^{n} \mathsf{P}(\mathrm{feature}_i), \qquad \mathsf{P}(\mathrm{observation} \mid \mathrm{class}) = \prod_{i=1}^{n} \mathsf{P}(\mathrm{feature}_i \mid \mathrm{class}),$$
where $n$ is the total number of features describing a given observation
(*For example, if an observation is presented as a sentence, then each
word can be a feature*). Thus,
$\mathsf{P}(\mathrm{class}|\mathrm{observation})$ now can be calculated
as

$$\mathsf{P}(\mathrm{class} \mid \mathrm{\mathrm{observation}}) = \mathsf{P}(\mathrm{class})\times \prod_{i=1}^{n}\frac{\mathsf{P}(\mathrm{feature}_i\mid \mathrm{class})}{\mathsf{P}(\mathrm{feature}_i)}\tag{1}$$

All the terms on the right-hand side can be estimated as corresponding
relative frequencies using available data\

**See [*this
link*](https://www.javatpoint.com/machine-learning-naive-bayes-classifier)
for more detailed explanations & examples :) Also you can watch [*this
video*](https://youtu.be/O2L2Uv9pdDA?si=-ohkHVDuu3sLLGMq) for more
examples!**

## Data description

There are 5 datasets uploaded on the cms (data.zip)

To determine your variant, take your team number from the list of teams
on cms and take *mod 5* - this is the number of your data set.

-   **0 - authors** This data set consists of citations of three famous
    writers: Edgar Alan Poe, Mary Wollstonecraft Shelley and HP
    Lovecraft. The task with this data set is to classify a piece of
    text with the author who was more likely to write it.

-   **1 - discrimination** This data set consists of tweets that have
    discriminatory (sexism or racism) messages or of tweets that are of
    neutral mood. The task is to determine whether a given tweet has
    discriminatory mood or does not.

-   **2 - fake news** This data set contains data of American news: a
    headline and an abstract of the article. Each piece of news is
    classified as fake or credible. The task is to classify the news
    from test.csv as credible or fake.

-   **3 - sentiment** All the text messages contained in this data set
    are labeled with three sentiments: positive, neutral or negative.
    The task is to classify some text message as the one of positive
    mood, negative or neutral.

-   **4 - spam** This last data set contains SMS messages classified as
    spam or non-spam (ham in the data set). The task is to determine
    whether a given message is spam or non-spam.

Each data set consists of two files: *train.csv* and *test.csv*. The
first one is used to find the probabilities of the corresponding classes
and the second one is used to test your classifier afterwards. Note that
it is crucial to randomly split your data into training and testing
parts to test the classifierʼs possibilities on the unseen data.

```{r}
# here goes a list of recommended libraries,
# though you may install other ones if they are needed
packages <- c("reshape2")
library(tidytext)
library(readr)
library(stringr)
library(tidyr)
library(dplyr)
library(ggplot2)
```

## Outline of the work

1.  **Data pre-processing** (includes removing punctuation marks and
    stop words, representing each message as a bag-of-words)
2.  **Data visualization** (it's time to plot your data!)
3.  **Classifier implementation** (using the training set, calculate all
    the conditional probabilities in formula (1))
4.  **Measurements of effectiveness of your classifier** (use the
    results from the previous step to predict classes for messages in
    the testing set and measure the accuracy, precision and recall, F1
    score metric etc)
5.  **Conclusions**

*!! do not forget to submit both the (compiled) Rmd source file and the
.html output !!*

## Data pre-processing

-   Read the *.csv* data files.
-   Сlear your data from punctuation or other unneeded symbols.
-   Clear you data from stop words. You don't want words as is, and, or
    etc. to affect your probabilities distributions, so it is a wise
    decision to get rid of them. Find list of stop words in the cms
    under the lab task.
-   Represent each test message as its bag-of-words. Here:
    <https://machinelearningmastery.com/gentle-introduction-bag-words-model/>
    you can find general introduction to the bag-of-words model and
    examples on to create it.
-   It is highly recommended to get familiar with R dataframes, it would
    make the work much easier to do.
-   Useful links:
    -   <https://steviep42.github.io/webscraping/book/bagofwords.html#tidytext> -
        example of using *tidytext* to count frequencies of the words.
    -   Basics of Text Mining in R:
        <http://rstudio-pubs-static.s3.amazonaws.com/256588_57b585da6c054349825cba46685d8464.html>
        . Note that it also includes an example on how to create a bag
        of words from your text document.

```{r}
list.files(getwd())
list.files("data/0-authors")
```

```{r}
test_path <- "test.csv"
train_path <- "train.csv"

stop_words <- read_file("stop_words.txt")
# https://stackoverflow.com/questions/27195912/why-does-strsplit-return-a-list
splitted_stop_words <- strsplit(stop_words, split='\n')
splitted_stop_words <- splitted_stop_words[[1]]
```

```{r}
train <-  read.csv(file = train_path, stringsAsFactors = FALSE)
test <-  read.csv(file = test_path, stringsAsFactors = FALSE)
```

```{r}
tidy_train <- train %>%
  mutate(doc_id = row_number()) %>%
  mutate(tweet = str_replace_all(tweet, "[^a-zA-Z\\s]", " ")) %>%
  unnest_tokens(word, tweet, token = "words") %>%
  filter(!word %in% splitted_stop_words) %>%
  filter(!is.na(word))

bag_of_words_train <- tidy_train %>% 
  count(doc_id, word) %>%
  pivot_wider(names_from = word, values_from = n, values_fill = 0)

X_train <- bag_of_words_train %>% select(-doc_id)
y_train <- train %>%
  mutate(doc_id = row_number()) %>%
  filter(doc_id %in% bag_of_words_train$doc_id) %>%
  pull(label)

tidy_test <- test %>%
  mutate(doc_id = row_number()) %>%
  mutate(tweet = str_replace_all(tweet, "[^a-zA-Z\\s]", " ")) %>%
  unnest_tokens(word, tweet, token = "words") %>%
  filter(!word %in% splitted_stop_words) %>%
  filter(!is.na(word))

bag_of_words_test <- tidy_test %>% 
  count(doc_id, word) %>%
  pivot_wider(names_from = word, values_from = n, values_fill = 0)

X_test <- bag_of_words_test %>% select(-doc_id)
y_test <- test$label

model_fixed <- SimpleNaiveBayes$new()
model_fixed$fit(X_train, y_train)
results <- model_fixed$score(test$tweet, y_test)
print(results)
```

## Data visualization

```{r}
garbage_words <- c("s", "t", "amp", "m")

tidy_train_clean <- tidy_train %>%
  filter(!is.na(word)) %>%
  filter(word != "user") %>%
  filter(!(word) %in% garbage_words)

top_words <- tidy_train_clean %>%
  group_by(label, word) %>%
  summarise(count = n(), .groups = "drop") %>%
  arrange(label, desc(count)) %>%
  group_by(label) %>%
  slice_max(order_by = count, n = 10)


ggplot(top_words, aes(x = reorder_within(word, count, label), 
                      y = count, fill = label)) +
  geom_col(show.legend = TRUE) +
  facet_wrap(~label, scales = "free_y") +
  coord_flip() +
  scale_x_reordered() + 
  labs(title = "Топ-10 word for each categoty",
       x = "Word",
       y = "Quantity") +
  theme_minimal()
```

Each time you work with some data, you need to understand it before you
start processing it. R has very powerful tools to make nice plots and
visualization. Show what are the most common words for negative and
positive examples as a histogram, word cloud etc. Be creative!

## Classifier implementation

```{r}
naiveBayes <- setRefClass("SimpleNaiveBayes",
  fields = list(
    class_probs = "numeric",
    feature_probs = "list",
    vocab = "character",
    classes = "character"
  ),
  methods = list(
    
    fit = function(X, y) {
      X_mat <- as.matrix(X)
      y_vec <- as.character(y)
      classes <<- unique(y_vec)
      class_counts <- table(y_vec)
      class_probs <<- as.numeric(class_counts / length(y_vec))
      names(class_probs) <<- classes
      vocab <<- colnames(X)
      feature_probs <<- list()
      
      for(cls in classes) {
        cls_indices <- which(y_vec == cls)
        
        if(length(cls_indices) > 0) {
          cls_data <- X_mat[cls_indices, , drop = FALSE]
          word_sums <- colSums(cls_data, na.rm = TRUE)
          total_words <- sum(word_sums)
          smoothed_probs <- (word_sums + 1) / (total_words + length(vocab))
          names(smoothed_probs) <- vocab
          
          feature_probs[[cls]] <<- smoothed_probs
        } else {
          uniform_probs <- rep(1/length(vocab), length(vocab))
          names(uniform_probs) <- vocab
          feature_probs[[cls]] <<- uniform_probs
        }
      }
    },
    
    predict = function(message) {
      message_clean <- tolower(gsub("[^a-zA-Z\\s]", " ", message))
      words <- strsplit(message_clean, "\\s+")[[1]]
      words <- words[words != "" & !words %in% splitted_stop_words]
      
      if(length(words) == 0) {
        return(classes[which.max(class_probs)])
      }
      log_probs <- sapply(classes, function(cls) {
        log_prob <- log(class_probs[cls])
        
        for(word in words) {
          if(word %in% vocab) {
            word_prob <- feature_probs[[cls]][word]
            if(word_prob > 0) {
              log_prob <- log_prob + log(word_prob)
            }
          }
        }
        return(log_prob)
      })
      return(classes[which.max(log_probs)])
    },
    
    score = function(texts, labels) {
      preds <- character(length(texts))
      
      for(i in 1:length(texts)) {
        preds[i] <- .self$predict(texts[i])
      }
      all_classes <- unique(c(labels, preds))
      conf_mat <- table(factor(preds, levels = all_classes), 
                       factor(labels, levels = all_classes))
      
      accuracy <- sum(diag(conf_mat)) / sum(conf_mat)
      precision <- recall <- f1 <- numeric(length(all_classes))
      names(precision) <- names(recall) <- names(f1) <- all_classes
      
      for(cls in all_classes) {
        TP <- conf_mat[cls, cls]
        FP <- sum(conf_mat[cls, ]) - TP
        FN <- sum(conf_mat[, cls]) - TP
        
        precision[cls] <- ifelse(TP + FP > 0, TP / (TP + FP), 0)
        recall[cls] <- ifelse(TP + FN > 0, TP / (TP + FN), 0)
        f1[cls] <- ifelse((precision[cls] + recall[cls]) > 0, 
                         2 * precision[cls] * recall[cls] / (precision[cls] + recall[cls]), 0)
      }
      
      return(list(
        accuracy = accuracy,
        precision = precision,
        recall = recall,
        f1 = f1,
        confusion_matrix = conf_mat
      ))
    }
  )
)
model <- naiveBayes$new()
model$fit(X_train, y_train)
results <- model$score(test$tweet, y_test)
print(results)

```

## Measure effectiveness of your classifier

-   Note that accuracy is not always a good metric for your classifier.
    Look at precision and recall curves, F1 score metric.

    When evaluating the model, it's important to understand the
    different types of classification results:

    -   A ***true positive*** result is one where the model correctly
        predicts the positive class.
    -   A ***true negative*** result is one where the model correctly
        predicts the negative class.
    -   A ***false positive*** result is one where the model incorrectly
        predicts the positive class when it is actually negative.
    -   A ***false negative*** result is one where the model incorrectly
        predicts the negative class when it is actually positive.

    Precision measures the proportion of true positive predictions among
    all positive predictions made by the model.

    $$
    Precision = \frac{TP}{TP+FP}
    $$

    Recall, on the other hand, measures the proportion of true positives
    identified out of all actual positive cases.

    $$
    Recall = \frac{TP}{TP+FN}
    $$

    F1 score is the harmonic mean of both precision and recall.

    $$
    F1 = \frac{2\times Precision \times Recall}{Precision + Recall}
    $$

    **See [this
    link](https://cohere.com/blog/classification-eval-metrics) to find
    more information about metrics.**

-   Visualize them.

-   Show failure cases.

```{r}
library(ggplot2)
conf_mat <- results$confusion_matrix
conf_df <- as.data.frame(as.table(conf_mat))
colnames(conf_df) <- c("Predicted", "Actual", "Freq")

ggplot(conf_df, aes(x = Actual, y = Predicted, fill = Freq)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Freq), color = "black", size = 4) +
  scale_fill_gradient(low = "white", high = "steelblue") +
  labs(title = "Confusion Matrix",
       x = "Actual",
       y = "Predicted") +
  theme_minimal()

visualize_metrics <- function(results) {
  metrics <- data.frame(
    Metric = c("Accuracy", 
               paste0("Precision (", names(results$precision), ")"),
               paste0("Recall (", names(results$recall), ")"),
               paste0("F1 (", names(results$f1), ")")),
    Value = c(results$accuracy, results$precision, results$recall, results$f1)
  )
  
  p1 <- ggplot(metrics, aes(x = Metric, y = Value, fill = Metric)) +
    geom_bar(stat = "identity") +
    geom_text(aes(label = round(Value, 3)), vjust = -0.5) +
    ylim(0,1) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    ggtitle("Основні метрики класифікації")
  
  print(p1)}
visualize_metrics(results)
```

## Conclusions

Overall, our Naive Bayes class, as expected, has three main functions.
The first calculates the probability of each word appearing in each of
the two classes. The second aggregates this data and computes the
probability of a given message belonging to each class. The third
evaluates overall performance based on the provided test .csv file,
essentially analyzing how well the method predicts.

The core mechanism for calculating the probability of each word is
Bayes’ theorem, while for entire sentences we simply use the average
probability across all words.

The main advantage of this approach is its speed when working with large
datasets.

Unfortunately, this method is quite sensitive: if one class has
significantly fewer words than the other, the likelihood of the program
assigning any message to that class drops sharply.

In conclusion, Naive Bayes provides a simple yet efficient baseline
model, but its effectiveness strongly depends on balanced and
representative training data.
